Item Construction Rules Revisited: Learnings from Measurement of Latent Variables with Gold Standard Items

### 2025-07-17 manuscript.qmd
update model description in methods section; add links to osf project and quarto website

author-note:
  disclosures:
    data-sharing: This is version 1 (last modification {{< meta date >}}) of our preprint published at <TODOK>. All materials (reproducible manuscript, analysis code, datasets, codebooks, questionnaires) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/p7492/>. A Quarto Manuscripts website is hosted at <TODO>. The data has previously been analyzed in the dissertation thesis of the first author at <https://doi.org/10.5282/edoc.23628>. As part of the dissertation, a preregistration was published at <https://osf.io/cz3uv/?view_only=9cac02db231b48629fea9ae53c3038b9> that differs from the analyses reported in the current manuscript.
    conflict-of-interest: The authors declare that there were no conflicts of interest concerning the authorship or the publication of this article.
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
language:
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: "This article investigates whether gold-standard items are helpful for questionnaire construction. Based on cognitive interviews (n = `r nrow(dat_interviews)`) and a pilot study (n = `r nrow(dat_pilot)`) with an initial item set, three 12-item scales were constructed to measure the physical traits of body height, body weight, and age. We collected data on these scales using response formats with either two (n = `r nrow(dat_dich)`) or six (n = `r nrow(dat_poly)`) categories. We also collected numeric self-reports of body weight, body height, and age as gold-standard items. Confirmatory factor analyses revealed that the gold-standard items did not consistently exhibit the highest loading on their corresponding latent variable. Furthermore, when controlling for the self-reported physical body height, body weight, and age as gold-standard items, as well as gender, we did not always find an interpretable, systematic residual variance. Finally, the pattern of correlations between the latent variables did not reflect the correlations between the self-reported gold-standard items, suggesting that the item scales and the gold-standard items do not have the same validity. While these results are consistent with previous studies, our analyses also showed that items with two response categories were at least as valid as those with six categories, contradicting past recommendations. When constructing a questionnaire, we would argue that the items intended to measure the latent variable most directly should have the highest loading on that variable. If this is not the case, the content validity is questionable at best. The implications are that intensive cognitive pretesting is necessary. Questionnaires with different response formats should be compared empirically and hypotheses about which items best represent the latent variable should be tested."
license: CC BY 4.0
bibliography: [references.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

@@

All materials (reproducible manuscript, analysis code, datasets, codebooks, questionnaires) for this manuscript are openly available on the Open Science Framework (OSF) at <https://osf.io/p7492/>. A Quarto Manuscripts website presenting the manuscript and electronic supplementary materials is hosted at <TODO>.

@@

First, we constructed items for the physical characteristics of body height, body weight, and age. Age was defined as chronological age or age in years, corresponding to the time elapsed since a person's birth [@montepareYouReOnly1989; @schwallDefiningAgeUsing2012]. Height was defined as the height of an upright person from the sole of the foot to the top of the head in centimeters, and body weight as the physical mass of a person in kilograms [@martinAnthropometrieAnleitungSelbstaendigen1929]. The initial items were developed by 124 psychology students [deductive item construction, @burischApproachesPersonalityInventory1984] and 24 persons [prototype approach, see @broughtonPrototypeStrategyConstruction1984] who deviated from the German population mean at least one standard deviation in the relevant characteristics (separately for women and men). The 24 persons were asked to think of prototypical behaviors for each construct to assess the whole spectrum of the latent variable. This process resulted in an initial set of 138 items (response categories were not tested). In a second step, we examined how these items were interpreted and whether they were connected to additional concepts unrelated to the constructs using cognitive interviews. We conducted interviews with a length of two to three hours and applied various cognitive procedures, including probing, paraphrasing, concurrent-think-aloud, and retrospective-think-aloud [@prueferVerfahrenEvaluationSurvey1996; @prueferKognitiveInterviews2005]. The literature on cognitive interviews recommends five to 30 interviews but indicates that the most serious problems can already be identified with a small number of interviews [@willisCognitiveInterviewing2005]. The interviews were conducted with `r nrow(dat_interviews)` people from the target group (`r nrow(dat_interviews[dat_interviews$Geschlecht == 1,])` women and `r nrow(dat_interviews[dat_interviews$Geschlecht == 2,])` men) aged between `r min(dat_interviews$Alter)` and `r max(dat_interviews$Alter)` (*M* = `r printnum(mean(dat_interviews$Alter))`, *SD* = `r printnum(sd(dat_interviews$Alter))`). The level of education ranged from high school diploma to university degree. After the cognitive interviews, misleading items were reformulated or eliminated [@faulbaumWasIstGute2009], resulting in 61 remaining items. In a third step, these 61 items were pilot tested online on a new sample, which initially consisted of 456 people. However, we excluded `r 456 - nrow(dat_pilot)` subjects because they had not completed the questionnaire, resulting in a sample of `r nrow(dat_pilot)` participants aged between `r min(dat_pilot$Alter)` and `r max(dat_pilot$Alter)` years (*M* = `r printnum(mean(dat_pilot$Alter))`, *SD* = `r printnum(sd(dat_pilot$Alter))`), including `r nrow(dat_pilot[dat_pilot$Geschlecht == 1,])` women (`r printnum(nrow(dat_pilot[dat_pilot$Geschlecht == 1,]) / nrow(dat_pilot) * 100)` %) and `r nrow(dat_pilot[dat_pilot$Geschlecht == 2,])` men (`r printnum(nrow(dat_pilot[dat_pilot$Geschlecht == 2,]) / nrow(dat_pilot) * 100)` %). The level of education was distributed as follows: `r round(nrow(dat_pilot[dat_pilot$Bildungsniveau == 1,]) / nrow(dat_pilot) * 100, 2)` % no school leaving certificate, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 2,]) / nrow(dat_pilot) * 100)` % secondary school leaving certificate/elementary school or equivalent, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 3,]) / nrow(dat_pilot) * 100)` % secondary school or equivalent, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 4,]) / nrow(dat_pilot) * 100)` % vocational baccalaureate or high school diploma, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 5,]) / nrow(dat_pilot) * 100)` % college degree or university degree, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 6,]) / nrow(dat_pilot) * 100)` % doctorate or habilitation. Out of all respondents, `r printnum(nrow(dat_pilot[dat_pilot$Muttersprache == 1,]) / nrow(dat_pilot) * 100)` % stated German as their native language. Based on these survey responses, we excluded items if a) their main item loading was not on the intended factor, b) their main and secondary loadings were almost equal, or c) they had loading below or equal to .30. Additionally, if (d) items had very similar content, items with lower loadings were excluded. More details on the process of item construction can be found in @eichhornItem2019. We reproduced the exploratory factor analyses, item total correlations and reliability estimates that were computed in the final item selection stage and report them in our electronic online materials (TODO). Based on these criteria 12 final items were selected for each physical trait (see @tbl-5).[^notewording] For each of the three questionnaires, two versions were created – one with a 6-point scale combining verbal endpoints ("does not apply" to "applies", only endpoints were labeled) with full numeric anchoring (numbers from 1 to 6) and one with a 2-point scale combining verbal labels ("does not apply" and "applies") with numeric anchors (numbers 0 and 1).

@@

All analyses were conducted in `r R.version$version.string`. A full list of software versions can be found in @apx-3. The data were analyzed using confirmatory factor analyses using the package *lavaan* [@R-lavaan]. We used the weighted least squares, mean and variance adjusted (WLSMV) estimator, treating the questionnaire items as ordinal and the self-reported physical height, weight, and age as continuous variables. First, we specified (1) a unidimensional model without any error correlations, and (2) a model with one latent variable and correlated errors. In both models the self-reported physical gold standard-items and the other 12 items were specified as indicators of a single latent variable. Second, to test whether there is a latent variable while controlling for the self-reported physical item and gender, (3) a model was specified with a single latent variable where the questionnaire items were additionally predicted by the self-reported physical item and gender. TODO. We also conducted reliability analyses using the package *MBESS* [@R-MBESS] and report McDonald's Omega as the reliability estimate of internal consistency.

>>>>

author-note:
  disclosures:
    data-sharing: This is version 1 (last modification {{< meta date >}}) of our preprint published on PsyArXiv. All materials (reproducible manuscript, analysis code, datasets, codebooks, questionnaires) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/p7492/>. A Quarto Manuscripts website is hosted at <https://florianpargent.github.io/gold-standard-items/>. The data has previously been analyzed in the dissertation thesis of the first author at <https://doi.org/10.5282/edoc.23628>. As part of the dissertation, a preregistration was published at <https://osf.io/cz3uv/?view_only=9cac02db231b48629fea9ae53c3038b9> that differs from the analyses reported in the current manuscript.
    conflict-of-interest: The authors declare that there were no conflicts of interest concerning the authorship or the publication of this article.
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
language:
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: "This article investigates whether gold-standard items are helpful for questionnaire construction. Based on cognitive interviews (n = `r nrow(dat_interviews)`) and a pilot study (n = `r nrow(dat_pilot)`) with an initial item set, three 12-item scales were constructed to measure the physical traits of body height, body weight, and age. We collected data on these scales using response formats with either two (n = `r nrow(dat_dich)`) or six (n = `r nrow(dat_poly)`) categories. We also collected numeric self-reports of body weight, body height, and age as gold-standard items. Confirmatory factor analyses revealed that the gold-standard items did not consistently exhibit the highest loading on their corresponding latent variable. Furthermore, when controlling for the self-reported physical body height, body weight, and age as gold-standard items, as well as gender, we did not always find an interpretable, systematic residual variance. Finally, the pattern of correlations between the latent variables did not reflect the correlations between the self-reported gold-standard items, suggesting that the item scales and the gold-standard items do not have the same validity. While these results are consistent with previous studies, our analyses also showed that items with two response categories were at least as valid as those with six categories, contradicting past recommendations. When constructing a questionnaire, we would argue that the items intended to measure the latent variable most directly should have the highest loading on that variable. If this is not the case, the content validity is questionable at best. The implications are that intensive cognitive pretesting is necessary. Questionnaires with different response formats should be compared empirically and hypotheses about which items best represent the latent variable should be tested."
license: CC BY 4.0
bibliography: [references.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

@@

All materials (reproducible manuscript, analysis code, datasets, codebooks, questionnaires) for this manuscript are openly available on the Open Science Framework (OSF) at <https://osf.io/p7492/>. A Quarto Manuscripts website presenting the manuscript and electronic supplementary materials is hosted at <https://florianpargent.github.io/gold-standard-items/>.

@@

First, we constructed items for the physical characteristics of body height, body weight, and age. Age was defined as chronological age or age in years, corresponding to the time elapsed since a person's birth [@montepareYouReOnly1989; @schwallDefiningAgeUsing2012]. Height was defined as the height of an upright person from the sole of the foot to the top of the head in centimeters, and body weight as the physical mass of a person in kilograms [@martinAnthropometrieAnleitungSelbstaendigen1929]. The initial items were developed by 124 psychology students [deductive item construction, @burischApproachesPersonalityInventory1984] and 24 persons [prototype approach, see @broughtonPrototypeStrategyConstruction1984] who deviated from the German population mean at least one standard deviation in the relevant characteristics (separately for women and men). The 24 persons were asked to think of prototypical behaviors for each construct to assess the whole spectrum of the latent variable. This process resulted in an initial set of 138 items (response categories were not tested). In a second step, we examined how these items were interpreted and whether they were connected to additional concepts unrelated to the constructs using cognitive interviews. We conducted interviews with a length of two to three hours and applied various cognitive procedures, including probing, paraphrasing, concurrent-think-aloud, and retrospective-think-aloud [@prueferVerfahrenEvaluationSurvey1996; @prueferKognitiveInterviews2005]. The literature on cognitive interviews recommends five to 30 interviews but indicates that the most serious problems can already be identified with a small number of interviews [@willisCognitiveInterviewing2005]. The interviews were conducted with `r nrow(dat_interviews)` people from the target group (`r nrow(dat_interviews[dat_interviews$Geschlecht == 1,])` women and `r nrow(dat_interviews[dat_interviews$Geschlecht == 2,])` men) aged between `r min(dat_interviews$Alter)` and `r max(dat_interviews$Alter)` (*M* = `r printnum(mean(dat_interviews$Alter))`, *SD* = `r printnum(sd(dat_interviews$Alter))`). The level of education ranged from high school diploma to university degree. After the cognitive interviews, misleading items were reformulated or eliminated [@faulbaumWasIstGute2009], resulting in 61 remaining items. In a third step, these 61 items were pilot tested online on a new sample, which initially consisted of 456 people. However, we excluded `r 456 - nrow(dat_pilot)` subjects because they had not completed the questionnaire, resulting in a sample of `r nrow(dat_pilot)` participants aged between `r min(dat_pilot$Alter)` and `r max(dat_pilot$Alter)` years (*M* = `r printnum(mean(dat_pilot$Alter))`, *SD* = `r printnum(sd(dat_pilot$Alter))`), including `r nrow(dat_pilot[dat_pilot$Geschlecht == 1,])` women (`r printnum(nrow(dat_pilot[dat_pilot$Geschlecht == 1,]) / nrow(dat_pilot) * 100)` %) and `r nrow(dat_pilot[dat_pilot$Geschlecht == 2,])` men (`r printnum(nrow(dat_pilot[dat_pilot$Geschlecht == 2,]) / nrow(dat_pilot) * 100)` %). The level of education was distributed as follows: `r round(nrow(dat_pilot[dat_pilot$Bildungsniveau == 1,]) / nrow(dat_pilot) * 100, 2)` % no school leaving certificate, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 2,]) / nrow(dat_pilot) * 100)` % secondary school leaving certificate/elementary school or equivalent, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 3,]) / nrow(dat_pilot) * 100)` % secondary school or equivalent, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 4,]) / nrow(dat_pilot) * 100)` % vocational baccalaureate or high school diploma, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 5,]) / nrow(dat_pilot) * 100)` % college degree or university degree, `r printnum(nrow(dat_pilot[dat_pilot$Bildungsniveau == 6,]) / nrow(dat_pilot) * 100)` % doctorate or habilitation. Out of all respondents, `r printnum(nrow(dat_pilot[dat_pilot$Muttersprache == 1,]) / nrow(dat_pilot) * 100)` % stated German as their native language. Based on these survey responses, we excluded items if a) their main item loading was not on the intended factor, b) their main and secondary loadings were almost equal, or c) they had loading below or equal to .30. Additionally, if (d) items had very similar content, items with lower loadings were excluded. More details on the process of item construction can be found in @eichhornItem2019. We reproduced the exploratory factor analyses, item total correlations and reliability estimates that were computed in the final item selection stage and report them in our electronic supplementary materials. Based on these criteria 12 final items were selected for each physical trait (see @tbl-5).[^notewording] For each of the three questionnaires, two versions were created – one with a 6-point scale combining verbal endpoints ("does not apply" to "applies", only endpoints were labeled) with full numeric anchoring (numbers from 1 to 6) and one with a 2-point scale combining verbal labels ("does not apply" and "applies") with numeric anchors (numbers 0 and 1).

@@

All analyses were conducted in `r R.version$version.string`. A full list of software versions can be found in @apx-3. The data were analyzed using confirmatory factor analyses using the package *lavaan* [@R-lavaan]. We used the weighted least squares, mean and variance adjusted (WLSMV) estimator, treating the questionnaire items as ordinal and the self-reported physical gold-standard items body height, body weight, and age as continuous variables. First, we specified (1) unidimensional models without any error correlations, and (2) models with one latent variable and correlated errors. In both models the gold-standard items and the other 12 items were specified as indicators of a single latent variable. Second, to test whether there is a latent variable while controlling for the gold-standard item and gender, (3) models were specified with a single latent variable where the questionnaire items were additionally predicted by the gold-standard item and gender. Third, to compare the correlations between the latent variables with the correlations between the gold-standard items, (4) full three factor models without correlated errors were estimated, in which only the questionnaire items load on their respective latent variable and correlations are estimated between the latent variables, between the gold-standard items, and between the latent variables and the gold-standard items. We also conducted reliability analyses using the package *MBESS* [@R-MBESS] and report McDonald's Omega as a reliability estimate of internal consistency for each scale.
